\Large\bf FlexibleSUSY --- A spectrum generator generator for supersymmetric models

### 2014-05-12 doc/paper.tex
improve the design goal and compilation sections a bit

In section \ref{Sec:Program} we describe the program in more detail
and explain our design goals.  In section \ref{Sec:download}
information on how to download and compile the code may be found along
with details on how to get started quickly.  In section
\ref{Sec:modfile} we describe how the user can create a new
FlexibleSUSY model file. A detailed description of the structure and
features of the generated code is then given in section
\ref{Sec:SpecGenStruct}.  In section \ref{Sec:Flexible} we describe
the various ways the code can be modified both at the meta code level
by writing model files and at the C++ code level by modifying the code
or adding new modules. Finally in section \ref{Sec:comparison} we
describe detailed comparisons between our generated code and existing
public spectrum generators as well as against the SPheno-like fortran
code which can be created using SARAH.

@@

  \fs makes use of the existing \sarah package
  \cite{Staub:2010ty,Staub:2009bi,Staub:2010jh,Staub:2012pb,Staub:2013tta}
  to obtain tree level expressions for the mass matrices and
  electroweak symmetry breaking conditions, one-loop self energies,
  one-loop tadpoles corrections and two-loop renormalisation group
  equations(RGEs) for the model specified in the model
  files. Additional corrections which have been calculated elsewhere,
  such as two-loop corrections to the Higgs masses\footnote{By default FlexibleSUSY has two loop corrections to the Higgs masses for the MSSM\cite{Degrassi:2001yf,Brignole:2001jy,Dedes:2002dy,Brignole:2002bz,Dedes:2003km}  and NMSSM \cite{Degrassi:2009yq} in FORTRAN files supplied by Pietro Slavich. These are the same corrections which are implemented in many of the public spectrum generators.} may be added in
  file specified by the user.

 \fs then
converts these algerbraic expressions from \sarah into C++ code using
the C++11 standard. \fs uses some parts of
SOFTSUSY\cite{Allanach:2001kg}, the very fast Eigen library
\cite{eigen}, as well as the gnu scientific library and the boost
library to create numerical routines which solve the RGEs and EWSB
conditions.  The main solver routine of \fs then calls these routines
to find simultaneous solutions for the EWSB, low energy data, and
other, user supplied, boundary conditions (e.g. the usual mSUGRA
universality conditions) and calculates the mass spectrum.

 The modular structure of the generated code allows for any individual
 components to be replaced with an alternative if available.  Indeed
 \fs has been intentionally designed to grow as alternative solvers
 and calculators are added and a subsequent release with an
 alternative lattice solver is already planned.  In addition if one
 already has an existing code for some other purpose and simply wishes
 to improve upon this by adding, e.g, new RGEs or self energies the
 modular framework makes this straightforward by adding or
 interfacing with the appropriate routine.

@@

\paragraph{Speed}
Exploring the parameter space of supersymmetric models with a high
number of free parameters is quite time consuming.  Therefore
FlexibleSUSY aims to produce spectrum generators with a short run-time.

The two most time consuming parts of a spectrum generator are the
calculation of the beta functions and the pole masses for mixed
particles.

The reason is the following: The RG solving algorithms usually need
$O(10)$ iterations to calculate a particle spectrum with $0.1\%$
precision.  During each iteration the Runge--Kutta algorithm needs to
calculate all beta functions $O(50)$ times.  Most two-loop beta
functions involve $O(50)$ matrix multiplications and additions.  All
together one arrives at $O(2.5\cdot 10^4)$ matrix multiplications.  To
optimize these matrix multiplications FlexibleSUSY uses the fast
linear algebra package \href{Eigen}{http://eigen.tuxfamily.org}, which
exploits C++ expression templates to make the matrix multiplications
easy to write and optimize.

The second most time consuming part is the precise calculation of the
pole masses for mixed particles.  For each particle $\psi_i$ the full
self-energy matrix $\Sigma^\psi_{ij}(p=m^\text{tree}_{\psi_i})$ has to
be evaluated.  Each self-energy matrix entry again involves the
evaluation of $O(50)$ Feynman diagrams, which involves the calculation
of vertices and loop functions.  All in all, one arrives at $O(500)$
Feynman diagram or $O(3\cdot 10^4)$ loop function evaluations.  To
speed up the calculation of the pole masses FlexibleSUSY makes use of
multi-threading, where each pole mass is calculated in a separate
thread.  By this technique one can gain a speed-up of $20$--$30\%$.

\paragraph{Modularity}
The large variety of supersymmetric models makes it likely that the
user needs to modify the generated spectrum generator source code.
FlexibleSUSY uses object orientation with C++ to modularize the source
code to make it easy for the user to modify and extend the spectrum
generator.  An important application here are the constraints: The RG
solvers provide a constraint interface,
\figref{fig:schematic-two-scale-constraint-interface},
%
\begin{figure}
  \centering
  \tikzumlset{fill class=white}
  \begin{tikzpicture}
    \umlclass[x=0, y=0]{RGFlow$<$Two\_scale$>$}{}{
      + solve()
    }
    \umlclass[x=6, y=0, type=abstract]{Constraint$<$Two\_scale$>$}{}{
      + \umlvirt{apply()}\\
      + \umlvirt{get\_scale() : double}
    }
    \umlclass[x=6, y=-3]{MyConstraint}{}{
      + apply()\\
      + get\_scale() : double
    }
    \umldep{RGFlow$<$Two\_scale$>$}{Constraint$<$Two\_scale$>$}
    \umlinherit{MyConstraint}{Constraint$<$Two\_scale$>$}
  \end{tikzpicture}
  \caption{Schematic two-scale constraint interface}
  \label{fig:schematic-two-scale-constraint-interface}
\end{figure}
%
which can be implemented to create the custom constraint
\code{MyConstraint}.  In the \code{apply()} function the constraint
application takes place and the \code{get_scale()} function specifies
the scale at which the constraint is to be applied.

\section{Download and compilation}
\label{Sec:download}
Flexible SUSY can be downloaded from \url{http://flexiblesusy.hepforge.org}.  
To use FlexibleSUSY please make sure you have the following installed.  
\begin{itemize}
\item Mathematica, version 7 or higher
\item SARAH, version 4.0.4 or higher \url{http://sarah.hepforge.org}
\item C++11 compatible compiler (g++ 4.4.7 or higher, clang++ 3.1 or
  higher)
\item Fortran compiler (gfortran, ifort etc.)
\item Eigen library, version 3.1 \url{http://eigen.tuxfamily.org}
\item Boost library, version 1.36.0 or higher
  \url{http://www.boost.org}
\item GNU scientific library \url{http://www.gnu.org/software/gsl/}
\item lapack (needed for the Lattice algorithm only)
  \url{http://www.netlib.org/lapack/}
\end{itemize}
%
Optional:
%
\begin{itemize}
\item Looptools, version 2.8 or higher
  \url{http://www.feynarts.de/looptools/}
\end{itemize}

\subsection{Quick Start}
Create a MSSM spectrum generator
  \begin{lstlisting}[language=bash]
$ ./createmodel --models=MSSM
$ ./configure --with-models=MSSM
$ make
  \end{lstlisting}
Run the spectrum generator using the default parameter point
  \begin{lstlisting}[language=bash]
$ ./models/MSSM/run_MSSM.x
  \end{lstlisting}
Run the spectrum generator using a SLHA input file
  \begin{lstlisting}[language=bash]
$ ./models/MSSM/run_MSSM.x --slha-input-file=templates/MSSM/LesHouches.in.MSSM --slha-output-file=LesHouches.out.MSSM
  \end{lstlisting}%% $

\section{Setting up a FlexibleSUSY model}
\label{Sec:modfile}
A \fs model is configured via a (Mathematica) model file, which in
case of the MSSM looks like this
%
\begin{lstlisting}[language=Mathematica]
FSModelName = "MSSM";

@@

All non-Standard Model input variables must be specified in the lists
\code{MINPAR} and \code{EXTPAR}.  Theses two variables refer to the
MINPAR and EXTPAR blocks in a SLHA input file \cite{Skands:2003cj}.
The list elements are two-component lists where the first entry is the
SLHA index in the MINPAR or EXTPAR blocks.  The second entry is the
name of the parameter.  The input parameters can be set to default
values in the \code{DefaultParameterPoint} list, see above.

Using the the variable \code{EWSBOutputParameters} the user can
specify the model parameters that are output of the electroweak
symmetry breaking conditions.  When doing EWSB, \fs will adjust these
parameters until the EWSB conditons are fulfilled.  In the CMSSM these
are \code{\\[Mu]} and \code{B[\\[Mu]]}.

Furthermore, the user has to specify the form of three model
constraints: low-scale, susy-scale and high-scale constraints.  In \fs
they are named as \code{LowScale}, \code{SUSYScale} and
\code{HighScale}.  For each constraint there is (i) a scale definition
(named after the constraint), (ii) an initial guess for the scale
(concatenation of the constraint name and \code{FirstGuess}) and (iii)
a list of parameter settings to be applied at the scale (concatenation
of the constraint name and \code{Input}).  Optional a minimum and a
maximum value for the scale can be given (concatenation of the
constraint name and \code{Minimum} or \code{Maximum}, respectively).
This avoids underflows or overflows of the scale value during the
iteration.  This is especially useful in models where the iteration is
very unstable and the value of the scale is very sensitive to the
model parameters.
%
\begin{itemize}
\item \emph{High-scale constraint:} The high-scale constraint is
  usuall the GUT-scale constraint, applied at the scale where the
  gauge couplings $g_1$ and $g_2$ unify.  The high-scale can be
  defined by an equation of the form \code{g1 == g2} or by fixed scale
  value.
\item \emph{Susy-scale constraint:} The susy-scale constraint is the
  constraint where \fs will impose the EWSB conditions and calculate
  the spectrum.  It is usually defined by
  $\sqrt{m_{\tilde{t}_1}m_{\tilde{t}_2}}$.
\item \emph{Low-scale constraint:} The low-scale constraint is the
  constraint where the gauge couplings $g_1$, $g_2$, $g_3$ and the
  Yukawa couplings $y_f$ will be calculated from given Standard Model
  input $\alpha_{\text{e.m.}}(M_Z)$, $\alpha_{s}(M_Z)$, $m_f^\pole$,
  $M_Z^\pole$, $M_W^\pole$, see
  Sec.~\ref{sec:calculation-of-gauge-couplings}--\ref{sec:calculation-of-yukawa-couplings}.
  This should be done at the $Z$ mass scale $M_Z$.  Note that the
  gauge and Yukawa couplings are calculated automatically and
  therefore do not need to be specified in the \code{LowScaleInput}
  variable.
\end{itemize}
%
Note, that \fs GUT-normalizes all gauge couplings.  Thus, the
high-scale definition takes the simple form \code{g1 == g2}.  Another
consequence is that in the calculation of the VEVs $v_u$ and $v_d$
from $M_Z$ and $\tan\beta$ at the low-scale the GUT-normalization has
to be taken into account, see above.

@@

To create a low-energy model one has to set
\code{OnlyLowEnergyFlexibleSUSY = True}.  In this case the high-scale
constraint will be ignored and only the low-scale and susy-scale
constraints are kept.  All model parameters that are not specified in
\code{MINPAR} or \code{EXTPAR} will be read from the corresponding
input blocks in the SLHA input file and will be set at the susy-scale.

\fs can create the helper function \code{get_lsp()}, which finds the
lightest supersymmetric particle (LSP).  To have this function be
created by \fs, the model file variable \code{PotentialLSPParticles}
must be set to a list of SUSY particles which are potential LSPs.  In
the model file example above, the particles \code{Chi}, \code{Cha},
\code{Glu}, \code{Sv}, \code{Su}, \code{Sd}, \code{Se} (neutralino,
chargino, gluino, sneutrino, up-type squark, down-type squark,
selectron) are considered to be LSPs.

@@

The parameters of the susy model together with their RGEs are stored
in the model class hierarchy, see the UML diagram in
\figref{fig:parameter-classes}.
%
\begin{figure}
  \centering
  \tikzumlset{fill class=white}
  \begin{tikzpicture}
    \umlclass[x=0, y=8, type=abstract]{Beta\_function}{
    }{
      + \umlvirt{get()}\\
      + \umlvirt{set()}\\
      + \umlvirt{beta()}\\
      + run\_to()
    }
    \umlclass[x=0, y=4]{<model>\_susy\_parameters}{
      -- susy parameters
    }{
      + get()\\
      + set()\\
      + beta()\\
      + run\_to()
    }
    \umlclass[x=0, y=0]{<model>\_soft\_parameters}{
      -- soft-breaking parameters
    }{
      + get()\\
      + set()\\
      + beta()\\
      + run\_to()
    }
    \umlinherit{<model>\_susy\_parameters}{Beta\_function}
    \umlinherit{<model>\_soft\_parameters}{<model>\_susy\_parameters}
  \end{tikzpicture}
  \caption{Model parameter class hierarchy.}
  \label{fig:parameter-classes}
\end{figure}

@@

All susy model parameters and their $\beta$-functions are contained in
the derived classes.  The structure of the $\beta$-functions of a
general susy model
\cite{Jones:1974pg,Jones:1983vk,West:1984dg,Martin:1993yx,Yamada:1993ga,MV94,Fonseca:2011vn,Sperling:2013eva,Sperling:2013xqa}
allows to split the parameters into two classes:
%
\begin{enumerate}
\item \emph{susy parameters:} gauge couplings, superpotential
  parameters and VEVs and
\item \emph{soft-breaking parameters} \cite{Girardello:1981wz}: soft
  linear scalar terms, soft bilinear scalar interactions, soft
  trilinear scalar interactions, soft gaugino mass terms and soft
  scalar squared masses.
\end{enumerate}
%
The $\beta$-functions of the susy parameters do only depend on the
susy parameters themselfs.  They do not depend on soft-breaking
parameters.  However, the $\beta$-functions of the soft-breaking
parameters depend on all model parameters in general.

This property is reflected in the C++ code as well: The class
\code{<model>_susy_parameters} direcly inherits from
\code{Beta_functions} and implements the $\beta$-functions of the susy
parameters.  The class of soft-breaking parameters
\code{<model>_soft_parameters} in turn inherits from
\code{<model>_susy_parameters} and implements the $\beta$-functions of
the soft-breaking parameters in terms of all model parameters.
This hierarchy allows to
%
\begin{enumerate}
\item use the RGE running of all model parameters via the common
  \code{Beta_function} interface and to
\item run the susy parameters independently of the soft-breaking
  parameters.
\end{enumerate}

\subsection{Boundary conditions}

The parameters of a susy model have to meet certain boundary
conditions.  For example, the running gauge and Yukawa couplings
should match the ones of the Standard Model at the scale $M_Z$.
Usually also high-scale boundary conditions on the soft-breaking
parameters are applied, as for example in mSUGRA scenarios.

@@

By default, \fs creates three default boundary conditions for each
susy model:
%
\begin{itemize}
\item The \emph{high-scale constraint} is indended to set boundary
  conditions on the model parameters at some very high scale, e.g.\
  the GUT scale $M_X$.  The high-scale is defined by the value given
  in the variable \code{HighScale}.  In the model file in
  \secref{Sec:modfile} it is defined to be the unification scale $M_X$
  where $g_1(M_X) = g_2(M_X)$.  The model parameters which, are set in
  the $apply()$ function at the high-scale, are extracted from the
  \code{HighScaleInput} variable.

\item The \emph{SUSY-scale constraint} is indended to set boundary
  conditions at the mass scale $M_S$ of the new susy particles.  The
  value of $M_S$ is defined by the model file variable
  \code{SUSYScale}.  In the example model file in \secref{Sec:modfile}
  it is defined to be $\sqrt{m_{\tilde{u}_1} m_{\tilde{u}_6}}$, where
  $m_{\tilde{u}_1}$ is the mass of the lightest, and $m_{\tilde{u}_6}$
  is the mass of the heaviest up-type squark.  The generated
  \code{apply()} function sets the model parameters, defined in
  \code{SUSYScaleInput}.  Finally, the \code{apply()} solves the EWSB
  equations at the loop level, by adjusting the parameters given in
  \code{EWSBOutputParameters} such that the effective Higgs potential
  is minimized.  See \secref{sec:ewsb} for a more detailed description
  of the EWSB in \fs.

\item The \emph{low-scale constraint} is indended to match the susy
  model to the Standard Model at the scale $M_Z$.  It does so by
  calculating the gauge and Yukawa couplings of the susy model from
  the known Standard Model fermion and gauge boson masses, as well as
  $\alpha_{\text{e.m.},\text{SM}}^{(5),\text{\MSbar}}(M_Z)$ and
  $\alpha_{\text{s},\text{SM}}^{(5),\text{\MSbar}}(M_Z)$.  See
  \secref{sec:calculation-of-gauge-couplings} and
  \ref{sec:calculation-of-yukawa-couplings} for more details.  In
  addition to the gauge and Yukawa couplings, the model parameter
  constraints given in \code{LowScaleInput} are imposed here.
\end{itemize}

@@

couplings $g_{i,\text{susy}}^{\text{\DRbar}}(M_Z)$ in the susy model

@@

and strong \DRbar\ couplings in the susy model

@@

$\theta_W$ in the susy model in the \DRbar\ scheme.  In the MSSM, for

@@

(GUT-normalized) $U(1)_Y$ and $SU(2)_L$ gauge couplings in the susy

@@

$y_f^{\text{\DRbar}}(M_Z)$ ($f=u,d,e$) in the susy model are

@@

\item At the $M_Z$ scale the susy model gauge couplings $g_{1,2,3}$

@@

\item The Yukawa couplings $y_{u,d,e}$ of the susy model are set to

@@

\item The susy parameters are run to the user-supplied first guess of

@@

\item Run model parameters to the susy-scale (\code{SUSYScale})

>>>>

In section \ref{Sec:Program} we describe the program in more detail
and explain our design goals.  In section \ref{Sec:download}
information on how to download and compile the code may be found along
with details on how to get started quickly.  In section
\ref{Sec:modfile} we describe how the user can create a new
FlexibleSUSY model file. A detailed description of the structure and
features of the generated code is then given in section
\ref{Sec:SpecGenStruct}.  In section \ref{Sec:Flexible} we describe
the various ways the code can be modified both at the meta code level
by writing model files and at the C++ code level by modifying the code
or adding new modules. Finally in section \ref{Sec:comparison} we
describe detailed comparisons between our generated code and existing
public spectrum generators as well as against the SPheno-like FORTRAN
code which can be created using SARAH.

@@

\fs makes use of the existing \sarah package
\cite{Staub:2010ty,Staub:2009bi,Staub:2010jh,Staub:2012pb,Staub:2013tta}
to obtain tree level expressions for the mass matrices and electroweak
symmetry breaking conditions, one-loop self energies, one-loop
tadpoles corrections and two-loop renormalisation group
equations(RGEs) for the model specified in the model files. Additional
corrections which have been calculated elsewhere, such as two-loop
corrections to the Higgs masses\footnote{By default FlexibleSUSY has
  two-loop corrections to the Higgs masses for the
  MSSM\cite{Degrassi:2001yf,Brignole:2001jy,Dedes:2002dy,Brignole:2002bz,Dedes:2003km}
  and NMSSM \cite{Degrassi:2009yq} in FORTRAN files supplied by Pietro
  Slavich. These are the same corrections which are implemented in
  many of the public spectrum generators.} may be added in file
specified by the user.

\fs then converts these algerbraic expressions from \sarah into C++
code using the C++11 standard. \fs uses some parts of
SOFTSUSY\cite{Allanach:2001kg}, the very fast Eigen library
\cite{eigen}, as well as the gnu scientific library and the boost
library to create numerical routines which solve the RGEs and EWSB
conditions.  The main solver routine of \fs then calls these routines
to find simultaneous solutions for the EWSB, low energy data, and
other, user supplied, boundary conditions (e.g.\ the usual mSUGRA
universality conditions) and calculates the mass spectrum.

The modular structure of the generated code allows for any individual
components to be replaced with an alternative if available.  Indeed
\fs has been intentionally designed to grow as alternative solvers and
calculators are added and a subsequent release with an alternative
lattice solver is already planned.  In addition if one already has an
existing code for some other purpose and simply wishes to improve upon
this by adding, e.g., new RGEs or self energies the modular framework
makes this straightforward by adding or interfacing with the
appropriate routine.

@@

\paragraph{Speed}

Exploring the parameter space of supersymmetric models with a high
number of free parameters is quite time consuming.  Therefore \fs aims
to produce spectrum generators with a short run-time.  The two most
time consuming parts of a SUSY spectrum generator are usually the
calculation of the two-loop $\beta$-functions and the pole masses of
mixed particles:
%
\begin{itemize}
\item \emph{Calculation of the $\beta$-functions:} The RG solving
  algorithms usually need $O(10)$ iterations between the high and the
  low scale to find a set of parameters consistent will all boundary
  conditions with a $0.01\%$ precision goal.  During each iteration
  the Runge-Kutta algorithm needs to calculate all $\beta$-functions
  $O(50)$ times.  Most two-loop $\beta$-functions involve $O(50)$
  matrix multiplications and additions.  All together one arrives at
  $O(10^4)$ matrix operations.  To optimize these, \fs uses the fast
  linear algebra package \href{Eigen}{http://eigen.tuxfamily.org}.
  Eigen uses C++ expression templates to remove temporary objects and
  enable lazy evaluation of the expressions.  It supports explicite
  vectorization, and provides fixed-size matrices to avoid dynamic
  memory allocation.  All of these features in combination result in
  very fast code for the calculation of the $\beta$-functions in \fs.
%
\item \emph{Calculation of the pole masses:} The second most time
  consuming part is the precise calculation of the pole masses for
  mixed particles.  For each particle $\psi_k$ in a multiplet the full
  self-energy matrix $\Sigma^\psi_{ij}(p=m^\text{tree}_{\psi_k})$ has
  to be evaluated.  Each self-energy matrix entry again involves the
  calculation of $O(50)$ Feynman diagrams, each involving the
  calculation of vertices and a loop-function.  All in all, one
  arrives at $O(500)$ Feynman diagrams and $O(10^4)$ loop function
  evaluations.  To speed up the calculation of the pole masses \fs
  makes use of multi-threading, where each pole mass is calculated in
  a separate thread.  This allows the operating system to distribute
  these calculations among different CPU cores.  With this technique
  one can gain a speed-up of $20$--$30\%$.
\end{itemize}

\paragraph{Modularity}
The large variety of supersymmetric models makes it likely that the
user wants to modify the generated spectrum generator source code.
\fs uses C++ object orientation features to modularize the source code
in order to make it easy for the user to modify, reuse and extend the
spectrum generator.  An important application of this concept are the
boundary conditions: The RG solver provides a common \code{Constraint}
interface, see \figref{fig:schematic-two-scale-constraint-interface}.
%
\begin{figure}
  \centering
  \tikzumlset{fill class=white}
  \begin{tikzpicture}
    \umlclass[x=0, y=0]{RGFlow<Two\_scale>}{}{
      + solve()
    }
    \umlclass[x=6, y=0, type=abstract]{Constraint<Two\_scale>}{}{
      + \umlvirt{apply()}\\
      + \umlvirt{get\_scale() : double}
    }
    \umlclass[x=6, y=-3]{MyConstraint}{}{
      + apply()\\
      + get\_scale() : double
    }
    \umldep{RGFlow<Two\_scale>}{Constraint<Two\_scale>}
    \umlinherit{MyConstraint}{Constraint<Two\_scale>}
  \end{tikzpicture}
  \caption{Schematic two-scale constraint interface}
  \label{fig:schematic-two-scale-constraint-interface}
\end{figure}
%
To create a custom constraint \code{MyConstraint}, the user implements
the interface by inheriting from \code{Constraint} and writing the
\code{apply()} and \code{get_scale()} methods (see
\secref{sec:boundary-conditions} for more details).  The so created
constraint class can then be passed to the RG solver to impose the
user-defined boundary condition at the chosen scale.

\section{Download and compilation}
\label{Sec:download}

\subsection{Requirements}

\fs can be downloaded from \url{http://flexiblesusy.hepforge.org}.  To
create a custom spectrum generator the following requirements are
necessary:
%
\begin{itemize}
\item MATHEMATICA, version 7 or higher
\item SARAH, version 4.0.4 or higher \url{http://sarah.hepforge.org}
\item C++11 compatible compiler (g++ 4.4.7 or higher, clang++ 3.1 or
  higher, icpc 12.1 or higher)
\item FORTRAN compiler (gfortran, ifort etc.)
\item Eigen library, version 3.1 or higher
  \url{http://eigen.tuxfamily.org}
\item Boost library, version 1.36.0 or higher
  \url{http://www.boost.org}
\item GNU scientific library \url{http://www.gnu.org/software/gsl/}
\item lapack \url{http://www.netlib.org/lapack/}
\end{itemize}
%
Optional:
%
\begin{itemize}
\item Looptools, version 2.8 or higher
  \url{http://www.feynarts.de/looptools/}
\end{itemize}

\subsection{Quick Start}

To create a MSSM spectrum generator the following three commands:
%
\begin{lstlisting}[language=bash]
$ ./createmodel --models=MSSM
$ ./configure --with-models=MSSM
$ make
\end{lstlisting}%% $
%
The first command creates the model directory \code{models/MSSM/}
together with a MSSM model file.  The \code{configure} scrip checks
the system requirements and creates the \code{Makefile}.  See
\code{./configure --help} for more options.  Executing \code{make}
will start MATHEMATICA to generate the spectrum generator and compile
it.  The resulting executable can be run like this:
\begin{lstlisting}[language=bash]
$ ./models/MSSM/run_MSSM.x
\end{lstlisting}%% $
%
An SLHA input file can be given via the \code{--slha-input-file=}
option, for example:
%
\begin{lstlisting}[language=bash]
$ ./models/MSSM/run_MSSM.x \
    --slha-input-file=templates/MSSM/LesHouches.in.MSSM
\end{lstlisting}%% $

\section{Setting up a FlexibleSUSY model}
\label{Sec:modfile}
A \fs model is configured via a (MATHEMATICA) model file
\code{FlexibleSUSY.m} in the model directory \code{models/<model>/}.
Many example model files can be found in \code{model_files/}.  The
default model file for the CMSSM looks like this:
%
\begin{lstlisting}[language=Mathematica]
FSModelName = "MSSM";

@@

All non-Standard Model input variables must be specified in the lists
\code{MINPAR} and \code{EXTPAR}.  Theses two variables refer to the
MINPAR and EXTPAR blocks in a SLHA input file \cite{Skands:2003cj}.
The list elements are two-component lists where the first entry is the
SLHA index in the MINPAR or EXTPAR blocks.  The second entry is the
name of the parameter.  The input parameters can be set to default
values in the \code{DefaultParameterPoint} list.

Using the the variable \code{EWSBOutputParameters} the user can
specify the model parameters that are output of the electroweak
symmetry breaking conditions.  When doing EWSB, \fs will adjust these
parameters until the EWSB conditons are fulfilled.  In the CMSSM these
are the superpotential parameter \code{\\[Mu]} and its corresponding
soft-breaking parameter \code{B[\\[Mu]]}.

Furthermore, the user has to specify the form of three model
constraints: low-scale, SUSY-scale and high-scale constraints.  In \fs
they are named as \code{LowScale}, \code{SUSYScale} and
\code{HighScale}.  For each constraint there is (i) a scale definition
(named after the constraint), (ii) an initial guess for the scale
(concatenation of the constraint name and \code{FirstGuess}) and (iii)
a list of parameter settings to be applied at the scale (concatenation
of the constraint name and \code{Input}).  Optional a minimum and a
maximum value for the scale can be given (concatenation of the
constraint name and \code{Minimum} or \code{Maximum}, respectively).
The latter can avoid underflows or overflows of the scale value during
the iteration.  This is especially useful in models where the
iteration is very unstable and the value of the scale is very
sensitive to the model parameters.  The meaning of the three
constraints is the following:
%
\begin{itemize}
\item \emph{High-scale constraint:} The high-scale constraint is
  usually the GUT-scale constraint, applied at the scale where the
  gauge couplings $g_1$ and $g_2$ unify.  The high-scale can be
  defined by an equation of the form \code{g1 == g2} or by fixed scale
  value.
%
\item \emph{SUSY-scale constraint:} The SUSY-scale constraint is the
  constraint where \fs imposes the EWSB conditions and calculate the
  pole mass spectrum.  Its scale $M_S$ is usually defined as
  $\sqrt{m_{\tilde{t}_1}m_{\tilde{t}_2}}$.  In the example above it
  has the value $M_S = \sqrt{m_{\tilde{u}_1}m_{\tilde{u}_6}}$, where
  $m_{\tilde{u}_1}$ and $m_{\tilde{u}_6}$ are the \DRbar\ masses of
  the lightest and heaviest up-type squark, respectively.
%
\item \emph{Low-scale constraint:} The low-scale constraint is the
  constraint where the gauge couplings $g_1$, $g_2$, $g_3$ and the
  Yukawa couplings $y_f$ are calculated from given Standard Model
  input $\alpha_{\text{e.m.}}(M_Z)$, $\alpha_{s}(M_Z)$, $m_f^\pole$,
  $M_Z^\pole$, $M_W^\pole$, see
  Sec.~\ref{sec:calculation-of-gauge-couplings}--\ref{sec:calculation-of-yukawa-couplings}.
  This should be done at the $Z$ mass scale $M_Z$.  The gauge
  couplings are calculated automatically and therefore do not need to
  be specified in the \code{LowScaleInput} variable.
\end{itemize}
%
\fs GUT-normalizes all gauge couplings.  Thus, the high-scale
definition takes the simple form \code{g1 == g2}.  Another consequence
is that in the calculation of the VEVs $v_u$ and $v_d$ from $M_Z$ and
$\tan\beta$ at the low-scale the GUT-normalization has to be taken
into account, see the example above.

@@

To create a low-energy model one has to set
\code{OnlyLowEnergyFlexibleSUSY = True}.  In this case the high-scale
constraint will be ignored and only the low-scale and SUSY-scale
constraints are kept.  All model parameters that are not specified in
\code{MINPAR} or \code{EXTPAR} will be read from the corresponding
input blocks in the SLHA input file and will be set at the SUSY-scale.

\fs can create the helper function \code{get_lsp()}, which finds the
lightest supersymmetric particle (LSP).  To have this function be
created the model file variable \code{PotentialLSPParticles} must be
set to a list of SUSY particles which are potential LSPs.  In the
model file example above, the particles \code{Chi}, \code{Cha},
\code{Glu}, \code{Sv}, \code{Su}, \code{Sd}, \code{Se} (neutralino,
chargino, gluino, sneutrino, up-type squark, down-type squark,
selectron) are considered to be LSPs.

@@

The parameters of the SUSY model together with their RGEs are stored
in the model class hierarchy, see the UML diagram in
\figref{fig:parameter-classes}.
%
\begin{figure}
  \centering
  \tikzumlset{fill class=white}
  \begin{tikzpicture}
    \umlclass[x=0, y=8, type=abstract]{Beta\_function}{
    }{
      + \umlvirt{get()}\\
      + \umlvirt{set()}\\
      + \umlvirt{beta()}\\
      + run\_to()
    }
    \umlclass[x=0, y=4]{<model>\_susy\_parameters}{
      -- susy parameters
    }{
      + get()\\
      + set()\\
      + beta()\\
      + run\_to()
    }
    \umlclass[x=0, y=0]{<model>\_soft\_parameters}{
      -- soft-breaking parameters
    }{
      + get()\\
      + set()\\
      + beta()\\
      + run\_to()
    }
    \umlinherit{<model>\_susy\_parameters}{Beta\_function}
    \umlinherit{<model>\_soft\_parameters}{<model>\_susy\_parameters}
  \end{tikzpicture}
  \caption{Model parameter class hierarchy.}
  \label{fig:parameter-classes}
\end{figure}

@@

All SUSY model parameters and their $\beta$-functions are contained in
the derived classes.  The structure of the $\beta$-functions of a
general SUSY model
\cite{Jones:1974pg,Jones:1983vk,West:1984dg,Martin:1993yx,Yamada:1993ga,MV94,Fonseca:2011vn,Sperling:2013eva,Sperling:2013xqa}
allows to split the parameters into two classes:
%
\begin{enumerate}
\item \emph{SUSY parameters:} gauge couplings, superpotential
  parameters and VEVs and
\item \emph{soft-breaking parameters} \cite{Girardello:1981wz}: soft
  linear scalar terms, soft bilinear scalar interactions, soft
  trilinear scalar interactions, soft gaugino mass terms and soft
  scalar squared masses.
\end{enumerate}
%
The $\beta$-functions of the SUSY parameters do only depend on the
SUSY parameters themselfs.  They do not depend on soft-breaking
parameters.  However, the $\beta$-functions of the soft-breaking
parameters depend on all model parameters in general.

This property is reflected in the C++ code as well: The class
\code{<model>_susy_parameters} direcly inherits from
\code{Beta_functions} and implements the $\beta$-functions of the SUSY
parameters.  The class of soft-breaking parameters
\code{<model>_soft_parameters} in turn inherits from
\code{<model>_susy_parameters} and implements the $\beta$-functions of
the soft-breaking parameters in terms of all model parameters.  The so
constructed class hierarchy allows to (i) use the RGE running of all
model parameters via the common \code{Beta_function} interface and to
(ii) run the SUSY parameters independently of the soft-breaking
parameters.

\subsection{Boundary conditions}
\label{sec:boundary-conditions}

The parameters of a SUSY model have to meet certain boundary
conditions.  For example, the running gauge and Yukawa couplings
should match the ones of the Standard Model at the scale $M_Z$.
Usually also high-scale boundary conditions on the soft-breaking
parameters are applied, as for example in mSUGRA scenarios.

@@

By default, \fs creates three default boundary conditions for each
SUSY model:
%
\begin{itemize}
\item The \emph{high-scale constraint} is indended to set boundary
  conditions on the model parameters at some very high scale, e.g.\
  the GUT scale $M_X$.  The high-scale is defined by the value given
  in the variable \code{HighScale}.  In the model file in
  \secref{Sec:modfile} it is defined to be the unification scale $M_X$
  where $g_1(M_X) = g_2(M_X)$.  The model parameters which, are set in
  the $apply()$ function at the high-scale, are extracted from the
  \code{HighScaleInput} variable.

\item The \emph{SUSY-scale constraint} is indended to set boundary
  conditions at the mass scale $M_S$ of the new SUSY particles.  The
  value of $M_S$ is defined by the model file variable
  \code{SUSYScale}.  In the example model file in \secref{Sec:modfile}
  it is defined to be $\sqrt{m_{\tilde{u}_1} m_{\tilde{u}_6}}$, where
  $m_{\tilde{u}_1}$ is the mass of the lightest, and $m_{\tilde{u}_6}$
  is the mass of the heaviest up-type squark.  The generated
  \code{apply()} function sets the model parameters, defined in
  \code{SUSYScaleInput}.  Finally, the \code{apply()} solves the EWSB
  equations at the loop level, by adjusting the parameters given in
  \code{EWSBOutputParameters} such that the effective Higgs potential
  is minimized.  See \secref{sec:ewsb} for a more detailed description
  of the EWSB in \fs.

\item The \emph{low-scale constraint} is indended to match the SUSY
  model to the Standard Model at the scale $M_Z$.  It does so by
  calculating the gauge and Yukawa couplings of the SUSY model from
  the known Standard Model fermion and gauge boson masses, as well as
  $\alpha_{\text{e.m.},\text{SM}}^{(5),\text{\MSbar}}(M_Z)$ and
  $\alpha_{\text{s},\text{SM}}^{(5),\text{\MSbar}}(M_Z)$.  See
  \secref{sec:calculation-of-gauge-couplings} and
  \ref{sec:calculation-of-yukawa-couplings} for more details.  In
  addition to the gauge and Yukawa couplings, the model parameter
  constraints given in \code{LowScaleInput} are imposed here.
\end{itemize}

@@

couplings $g_{i,\text{susy}}^{\text{\DRbar}}(M_Z)$ in the SUSY model

@@

and strong \DRbar\ couplings in the SUSY model

@@

$\theta_W$ in the SUSY model in the \DRbar\ scheme.  In the MSSM, for

@@

(GUT-normalized) $U(1)_Y$ and $SU(2)_L$ gauge couplings in the SUSY

@@

$y_f^{\text{\DRbar}}(M_Z)$ ($f=u,d,e$) in the SUSY model are

@@

\item At the $M_Z$ scale the SUSY model gauge couplings $g_{1,2,3}$

@@

\item The Yukawa couplings $y_{u,d,e}$ of the SUSY model are set to

@@

\item The SUSY parameters are run to the user-supplied first guess of

@@

\item Run model parameters to the SUSY-scale (\code{SUSYScale})
